---
title: "07_TNBC_2018_LDA_multiChains"
knit: (function(input, encoding) {
  rmarkdown::render(input = input,
                    output_dir = here::here("Output", "HTML"),
                    knit_root_dir = rprojroot::find_rstudio_root_file())})
output:
  html_document: 
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document: 
    toc: yes
date: "2023-02-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#knitr::opts_knit$set(root.dir = "../")
#knitr::opts_knit$set(output.dir = "../Output_Documents/")
options(dplyr.summarise.inform = FALSE)
```


## Library
```{r library, warning=FALSE, message=FALSE}
library(rstan)
library(dplyr)
library(ggplot2)
library(randomcoloR)
library(SpatialExperiment)
```


```{r plot theme}
theme_set(theme_minimal())
theme_update(
  text = element_text(size = 10),
  legend.text = element_text(size = 10)
)


# plot theme
remove_x <- theme(
  axis.text.x = element_blank(),
  axis.ticks.x = element_blank(),
  axis.title.x = element_blank()
)

remove_y <- theme(
  axis.text.y = element_blank(),
  axis.ticks.y = element_blank(),
  axis.title.y = element_blank()
)



# force not to use scientific notation
options(scipen=999)
```



## Dataset

Load the TNBC SPE object.
```{r dataset, warning=FALSE, message=FALSE}
load(here::here("Output", "Data", "03_SPE", "03_TNBC_2018_spe.rds"))
spe
```

## Pre-process
### Document-Term Matrix

We want to create the Document-Term matrix of the dataset where the rows are the sample identity and columns are the cell types fill with counts. The expected dimension of the DTM is 39 rows by 16 columns.

```{r dtm}
dtm <- table(spe$sample_id, spe$mm)
dim_names <- dimnames(dtm)
dim(dtm)
```


### Posterior Sampling

We want to estimate the parameters $\theta$ and $B = \left[\beta_{1}, \cdots, \beta_{K} \right]^{T}$, and we estimate using Hamiltonian Monte Carlo(HMC) No-U-Turn-Sampler(NUTS) with one chain and 500, 1000, and 2000 iterations. Out of these iterations, half of the iteration were used as warm-up samples.

We will apply `get_lda_result()` to perform LDA and save results.

```{r lda result, eval=FALSE}
source(here::here("Notebooks", 
                  "06_LDA_scripts", "06_get_lda_result.R"))

samples <- get_lda_result(K = 5, alpha = 0.8, gamma = 0.8,
                          dtm = dtm, iter = 2000, chains = 4,
                          file_fold = "07_LDA_multiChains", save_file = TRUE)
```

Load the existing LDA estimations.
```{r load result}
source(here::here("Notebooks", 
                  "06_LDA_scripts", "06_load_lda_result.R"))

file_fold = "07_LDA_multiChains"
lda_file = "JABES_Endo_all_K_5_ite_2000.RData"

samples <- load_lda_result(
  file = here::here("Output", "Data", 
                   file_fold, lda_file)
)
```

```{r}
K = 5
iter = 2000
chain = 4
stan.fit <- samples[[1]]
samples <- samples[[2]]
```

## Topics Analysis

### Theta

We want to extract estimation of parameters $\theta$ and $B = \left[\beta_{1}, \cdots, \beta_{K} \right]^{T}$. 

The dimension of $\theta$ is N x D x K, where k is the number of iterations with 39 patients and 3 topics. Each value is the probability of each patient in each topic in each iteration. Each value is the probability in each iteration of each topic of each tissue sample.

We are also interested in what cell types are in each topic, which is the $\beta$ estimation. The expected dimension is N x K x T, where $k$ is the number of iterations for 3 topics and 16 cell types. Each value is the probability in each iteration of each cell type in each topic.

Each value is the $k$ topic proportion in $d$ sample in $n$ iteration.
Each value is the $t$ term proportion in $k$ topic in $n$ iteration.

```{r parameter estimation}
# theta estimation
theta <- samples$theta
dim(theta)

# beta estimation
beta <- samples$beta
dim(beta)
```


### Alignment matrix



```{r}
source(here::here("Notebooks", 
                  "06_LDA_scripts", "06_alignmentMatrix.R"))

aligned_matrix <- alignmentMatrix(
  theta = theta,
  spe = spe,
  K = K,
  iter = iter,
  chain = chain,
  #cor_method = "cosine",
  SampleID_name = "sample_id"
)
```


### Align topic proportion

We align the topic proportion in each specimen across four chains.

```{r align-theta}
library(abind)
source(here::here("Notebooks", 
                  "06_LDA_scripts", "06_thetaAligned.R"))

theta_aligned <- thetaAligned(
  theta = theta, 
  K = K, 
  aligned = aligned_matrix, 
  iter = iter,
  chain = chain,
  )
```


### Align ASV proportion

We align the ASVs proportion in each topic across four chains.
```{r align-beta}
source(here::here("Notebooks", 
                  "06_LDA_scripts", "06_betaAligned.R"))


beta_aligned <- betaAligned(
  beta = beta, 
  K = K, 
  aligned = aligned_matrix, 
  iter = iter,
  chain = chain,
  ) 
```


## Visualization

### Plot topic proportion

Plot the topic proportion in specimens and we can save it for publication purposes. We can draw an informative summary of bacterial communities in each phenotype.

```{r plot-topic-prop, fig.width=20, fig.height=8}
source(here::here("Notebooks", 
                  "06_LDA_scripts", "06_plotTopicProportion.R"))

p_topicProp <- plotTopicProportion(
  spe = spe,
  theta_aligned = theta_aligned,
  K = K,
  iter = iter,
  chain = chain,
  SampleID_name = "sample_id"
)
p_topicProp

#ggsave(filename = "Topic_Prop_5.png", path="/Users/henzhwang/Desktop/",
#       width = 15, height = 10)
```


### Plot ASVs distribution

We plot the ASVs distribution in each topic.
```{r plot cellType Dist, fig.width=10, fig.height=10}
source(here::here("Notebooks", 
                  "06_LDA_scripts", "06_plotCellTypeDistribution.R"))

p_cellTypeDist <- plotCellTypeDistribution(
  spe = spe,
  K = K,
  beta_aligned = beta_aligned,
  col_names_beta_hat = c("iterations", "Topic", "Cell.Type", "beta_h")
)

p_cellTypeDist
#ggsave(filename = "cellType_dist_5.png", path="/Users/henzhwang/Desktop/",
#       width = 15, height = 10)
```

```{r}
# p <- ggpubr::ggarrange(
#   p_topicProp, p_cellTypeDist,
#   labels = "AUTO",
#   widths = 10, heights = c(4, 5),
#   nrow = 2
#   #common.legend = TRUE, lengend = "top"
# )
# 
# ggsave(filename = "topic_cellType_prop_plots.eps", device = "eps",
#        width = 12, height = 9,
#        path = here::here("Output", "Figures"))

```


## Diagnostic plot

Perform computation of effective sample size for each parameter
```{r}
source(here::here("Notebooks", 
                  "06_LDA_scripts", "06_diagnosticsPlot.R"))

p_diag_5_2000 <- diagnosticsPlot(
  theta_aligned = theta_aligned,
  beta_aligned = beta_aligned,
  iter = iter,
  chain = chain,
) 
```

```{r}
# bulk effective sample size
p_diag_5_2000[[1]]
# ggsave(filename = "ESS_5.png", path="/Users/henzhwang/Desktop/",
#        width = 8, height = 5)
```

```{r}
# rhat plots
p_diag_5_2000[[2]]
#ggsave(filename = "rhat_5.png", path="/Users/henzhwang/Desktop/",
#      width = 10, height = 6)
```


## Model Asseessement

We want to perform goodness of fit test using the posterior estimates and observed data.

```{r}
source(here::here("Notebooks", 
                  "06_LDA_scripts", "06_modelAssessment.R"))
#cellTypeIndexToPlot = c(2, 3, 5, 12, 15)

p_model_assessment <- modelAssessment(dtm = dtm,
                samples = samples,
                warm_up_iter = NULL,
                iter = iter,
                cellTypeIndexToPlot = c(1:16)
                )
p_model_assessment
```


```{r}
# ggpubr::ggarrange(
#   ggpubr::ggarrange(
#     p_diag_5_2000[[1]], p_diag_5_2000[[2]],
#     ncol = 2,
#     labels = c("A", "B")
#   ),
#   ggpubr::ggarrange(
#     p_model_assessment,
#     labels = c("C")
#   ),
#   heights = c(3, 5),
#   nrow = 2
# )
# 
# ggsave(filename = "diag_assess_plots.eps", device = "eps",
#        width = 10, height = 6,
#        path = here::here("Output", "Figures"))
```




## Simple function

We have also written above workflow into one simple function, which will return a list of the above output. If used chain in the posterior sampling is greater than one, the function will return a list of 

```{r}
# sourcing
source(here::here("Notebooks", 
                  "06_LDA_scripts", "06_LDA_analysis.R"))
```


```{r}
TNBC_2018_LDA_3_2000 <- LDA_analysis(
  spe = spe,
  SampleID_name = "sample_id",
  cellType_name = "mm",
  K = 3,
  alpha = 0.8,
  gamma = 0.8,
  iter = 2000,
  chain = 4,
  col_names_theta_all = c("iteration", "Sample", "Topic", "topic.dis"),
  col_names_beta_hat = c("iterations", "Topic", "Cell.Type", "beta_h"),
  stan_file_path = here::here("Notebooks", "06_LDA_scripts", "06_lda.stan"),
  load_file = TRUE,
  save_file = FALSE,
  file_default = TRUE,
  file_fold = "07_LDA_multiChains"
)
```


```{r}
TNBC_2018_LDA_5_2000 <- LDA_analysis(
  spe = spe,
  SampleID_name = "sample_id",
  cellType_name = "mm",
  K = 5,
  alpha = 0.8,
  gamma = 0.8,
  iter = 2000,
  chain = 4,
  col_names_theta_all = c("iteration", "Sample", "Topic", "topic.dis"),
  col_names_beta_hat = c("iterations", "Topic", "Cell.Type", "beta_h"),
  stan_file_path = here::here("Notebooks", "06_LDA_scripts", "06_lda.stan"),
  load_file = TRUE,
  save_file = FALSE,
  file_default = TRUE,
  file_fold = "07_LDA_multiChains"
)
```

```{r}
TNBC_2018_LDA_7_2000 <- LDA_analysis(
  spe = spe,
  SampleID_name = "sample_id",
  cellType_name = "mm",
  K = 7,
  alpha = 0.8,
  gamma = 0.8,
  iter = 2000,
  chain = 4,
  col_names_theta_all = c("iteration", "Sample", "Topic", "topic.dis"),
  col_names_beta_hat = c("iterations", "Topic", "Cell.Type", "beta_h"),
  stan_file_path = here::here("Notebooks", "06_LDA_scripts", "06_lda.stan"),
  load_file = TRUE, # load_file = FALSE
  save_file = FALSE, # save_file = TRUE
  file_default = TRUE,
  file_fold = "07_LDA_multiChains"
)
```



```{r fig.width=20, fig.height=15}
# topic proportions plots 
p_topic_prop_3_2000 <- TNBC_2018_LDA_3_2000[[9]]
p_topic_prop_5_2000 <- TNBC_2018_LDA_5_2000[[9]]
p_topic_prop_7_2000 <- TNBC_2018_LDA_7_2000[[9]]

# plot list
topic_prop_plot_list <- list(
  p_topic_prop_3_2000 + remove_x,
  p_topic_prop_5_2000 + remove_x,
  p_topic_prop_7_2000 
)

# plot
ggpubr::ggarrange(
  plotlist = topic_prop_plot_list,
  labels = c("A", "B", "C"),
  ncol = 1, nrow = 3
)


#p_topic_prop_3_2000 + p_topic_prop_5_2000 + p_topic_prop_8_2000

# p_topic_prop_3_2000
# p_topic_prop_5_2000
# p_topic_prop_8_2000
# ggsave(filename = "Topics.png", path="/Users/henzhwang/Desktop/",
#        width = 18, height = 15)
```


```{r}
# cell type proportions plots
p_cellType_prop_3_2000 <- TNBC_2018_LDA_3_2000[[10]]
p_cellType_prop_5_2000 <- TNBC_2018_LDA_5_2000[[10]]
p_cellType_prop_7_2000 <- TNBC_2018_LDA_7_2000[[10]]
```


```{r fig.width=15, fig.height=8}
# plot list
cellType_prop_plot_list <- list(
  p_cellType_prop_3_2000,
  p_cellType_prop_5_2000 + remove_y,
  p_cellType_prop_7_2000 + remove_y
)

# plot
ggpubr::ggarrange(
  plotlist = cellType_prop_plot_list,
  labels = c("A", "B", "C"),
  common.legend = TRUE, legend = "top",
  ncol = 3, nrow = 1
)


#p_cellType_prop_3_2000 + p_topic_prop_5_2000 + p_topic_prop_8_2000

# p_cellType_prop_3_2000
# p_cellType_prop_5_2000
# p_cellType_prop_8_2000

# ggsave(filename = "cellType.png", path="/Users/henzhwang/Desktop/",
#       width = 15, height = 12)
```

